\textbf{Appendix}\\ Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-08-21 Chapter_0/04_Manuscript/Manuscript_Backup_2020-02-04.tex
first commit

% Title of the document
{\textsc{ \LARGE Ranging Wide: Movement Corridors of Dispersing African Wild
Dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi Transfrontier Conservation
Area}}\\[0.4cm]

@@

\begin{abstract}
The African wild dog \textit{(Lycaon pictus)} is an endangered species that
currently only survives in a few scattered subpopulations in Sub-Saharan Africa.
Long-term viability of the species thus requires the identification and
preservation of key dispersal corridors that connect such subpopulations. We
collected GPS data of 16 dispersing coalitions originating from a free-ranging
population in northern Botswana and used integrated step selection analysis to
create a permeability surface spanning over the entire Kavango-Zambezi
Transfrontier Conservation Area (KAZA-TFCA), the largest transboundary
conservation area on earth. Based on our permeability surface, we calculated
least-cost paths and least-cost corridors between selected locations within the
KAZA-TFCA. Our results show that dispersal corridors typically run through open
shrubs/grassland and in the vicinity of water. Corridors rarely cross
waterbodies, areas with high human presence, or areas densely vegetated by
trees. Several dispersal corridors meet in the Okavango-Linyanty ecosystem,
which highlights the importance of northern Botswana for the conservation of
this charismatic species.
\end{abstract}
\newpage

@@

\section{Introduction}
% Importance of Connectivity
Successful conservation of fragmented and isolated subpopulations requires
maintenance and enhancement of connectivity between habitats occupied by these
subpopulations \citep{Hanski.1998}. In this context, connectivity describes the
degree to which the landscape matrix facilitates or impedes movement of
individuals and thereby genes of the focal species between habitat patches
\citep{Taylor.1993, Clobert.2012}. Thus, sufficient connectivity facilitates
dispersal, which in turn promotes genetic exchange, reinforcement and rescuing
of small populations, and colonization or recolonization of unoccupied habitats
\citep{Brown.1977, Hanski.1998, MacArthur.2001, Frankham.2002, Leigh.2012}. All
these processes crucially contribute to long-term population viability, which
emphasizes the importance of dispersal and connectivity. However, increasing
anthropogenic pressure and ongoing habitat loss and fragmentation reduce
connectivity on a worldwide scale \citep{Fahrig.2003, Barnosky.2012}. As a
result, management actions for the conservation of endangered species are
increasingly aimed towards preserving and restoring connectivity through the
identification and preservation of key dispersal corridors that link isolated
habitats \citep{Fahrig.2003, Heller.2009}. An unprecedented initiative to
restore connectivity on a large scale is represented by the creation of the
Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), a protected area
which traverses five African countries and spans over 520,000
km\textsuperscript{2}. While such initiatives are expected to bring benefits to
wide-ranging species such as elephants (\textit{Loxodonta africana}), African
wild dogs (\textit{Lycaon pictus}), and lions (\textit{Panthera leo}), the
extent to which they are effective for targeted species and whether boundaries
reflect natural dispersal corridors is rarely assessed.

% Connectivity Assessment
In recent years, there has been a growing body of literature that uses animal
movement data to identify dispersal corridors and to assess connectivity on
large scales \citep{Chetkiewicz.2006}. Identification of potential dispersal
corridors between suitable habitat patches traditionally relies on the
estimation of a permeability surface (elsewhere referred to as resistance
surface, when inverted), which renders the ease or willingness at which the
focal species traverses a specific landscape. Latest empirical methods to
predict landscape permeability based on recorded GPS relocation data include
\textit{resource selection functions} (RSF, \cite{Boyce.2002}), \textit{step
selection functions} (SSF, \cite{Fortin.2005}), and \textit{path selection
functions} (PSF, \cite{Cushman.2010}). All methods have in common that they
infer habitat preferences of the focal species by comparing covariates (e.g.
environmental and anthropogenic) at visited locations to the same covariates at
randomly selected locations that are deemed available but presumably unused by
the animal. Based on inferred habitat preferences the methods further assume
that habitats preferred by the focal species provide high permeability, whereas
avoided habitats provide low permeability. Regardless of the method chosen (i.e.
RSF, SSF, or PSF), all approaches require adequate GPS relocation data that is
representative for the process being studied. More specifically, it has been
shown that GPS relocation data collected on dispersing individuals is better
suited for detecting dispersal corridors than GPS relocation data collected on
resident individuals \citep{Elliot.2014}. Few studies have, however, partitioned
their GPS relocation data according to behavioral mode such as \textit{resident}
vs. \textit{dispersing} or \textit{active} vs. \textit{inactive} periods of
movement \citep{Wilson.2012, Vasudev.2015}. As a consequence, most permeability
surfaces upon which dispersal corridors are identified are created using GPS
relocation data that was collected on resident individuals. This introduces
severe biases and substantially reduces the power to reveal meaningful
corridors, for dispersing individuals have different needs and motivations
compared to resident individuals \citep{Killeen.2014, Elliot.2014, Cozzi.2020}.
For instance, \cite{Elliot.2014} demonstrated marked differences in predicted
permeability surfaces for dispersing lions in the KAZA-TFCA depending on whether
permeability was assessed using data collected during residence or dispersal.
\cite{Jackson.2016} made similar observations for African wild dogs and revealed
higher permeability for dispersers than for residents. Moreover,
\cite{Abrahms.2017} found a significantly greater overlap between predicted and
observed corridors for three dispersing wild dogs when predictions were based on
information collected only during wild dog activity, rather than across the
entire day.

% African wild dogs
A highly endangered flagship species for conservation efforts across the
KAZA-TFCA, but for which systematic GPS relocation data on dispersing
individuals is largely unavailable, is the African wild dog
\citep{VanDerMeer.2016}. While once present across the entire Sub-Saharan
continent, the species has been widely extirpated through human persecution,
habitat destruction, and disease outbreaks \citep{Woodroffe.2012}. As a result,
the African wild dog has become one of Africa's most endangered large carnivores
and currently only survives in small, spatially scattered subpopulations
\citep{Woodroffe.2012}. Within these subpopulations, wild dogs form cooperative
breeding groups of up to thirty individuals in which a single dominant pair
monopolizes the majority of reproduction \citep{Frame.1979, Fuller.1992,
Creel.2002}, while subordinates of both sexes help to feed newborn pups
\citep{Malcolm.1982}. After reaching 15 months of age, male and female offspring
disperse from their natal pack in same-sex coalitions and search for unrelated
mates and a suitable territory to settle (\citeauthor{McNutt.1996}, 1996;
\citeauthor{Behr.2019}, unpublished). During such dispersal events, wild dogs
are able to cover several hundred kilometers, sometimes crossing risky areas
that are strongly influenced by humans \citep{DaviesMostert.2012, Masenga.2016,
Cozzi.2020}. Despite the importance of dispersal for the remaining wild dog
subpopulations, little is known about the corridors of dispersing individuals
and connectivity between subpopulations in the KAZA-TFCA in general.

% Covariate Motivation
Wild dog movements and spatial behavior are known to be influenced by a suite of
environmental and anthropogenic factors. For instance, \cite{Abrahms.2017} found
that wild dog movements during residence mainly depended on land cover,
protection status and anthropogenic features. Similar observations on resident
packs were made by \cite{Pretorius.2019}, who's results indicate that pack
presence was contingent on factors such as agricultural activity, road presence,
and protection status of the area. With respect to dispersing individuals
\cite{Whittington.2011} demonstrated that elevation, land cover, road density,
and human density best explained the distribution of recorded sightings in South
Africa. Also based on data from dispersing wild dogs, \cite{Jackson.2016} showed
that land cover, human density and road density significantly influenced the
probability of occurrence of dispersing wild dogs. Finally, \cite{Masenga.2016}
reported a low willingness of dispersers to cross human-dominated landscapes.
Large scale spatial information on the above mentioned factors can be obtained
either through freely available online services or through remote sensed
satellite imagery.

% Our Study
We used GPS relocation data collected on 16 wild dogs in as many dispersing
coalitions from a free ranging population in northern Botswana to estimate the
degree of selection or avoidance for environmental and anthropogenic landscape
features. We then used the obtained information in combination with relevant
spatial covariates to create a permeability surface for the entire KAZA-TFCA.
Lastly, we used the permeability surface as input for the computation of
least-cost paths and least-cost corridors between selected locations. This
allowed us to highlight key dispersal corridors within and outside the
boundaries of KAZA-TFCA and concluded with the identification of a ``dispersal
hub'' and of dispersal corridors that run outside formally protected areas.

\newpage
\section{Methods}
\subsection{Study Area}
% Both Study Areas
We differentiated between a \textit{core study area}, where all dispersers
originated from, and an \textit{extended study area}, which was represented by a
bounding box encompassing the KAZA-TFCA \citep{Cozzi.2020} and for which we
calculated connectivity.

% Core Study Area
The \textit{core study area} was located in the eastern section of the Okavango
Delta in northern Botswana (centered at -19\degree 25'30''S, 23\degree 37'30''E;
elevation ca. 950 m, orange rectangle in \Cref{StudyArea}b) and comprised parts
of the Moremi Game Reserve and adjacent wildlife management areas. The Okavango
Delta is one of the main strongholds of African wild dogs in southern Africa and
likely acts as a source for the recolonization of surrounding habitats
\citep{Woodroffe.2012, Cozzi.2013}. It is the earth's largest inland delta and
is characterized by diverse habitats, ranging from permanent river systems,
mopane woodland, and acacia forests, to open grasslands and seasonal swamps
\citep{Broekhuis.2013}. Rainfall is seasonal and lasts from November to March,
whereas the remainder of the year remains dry \citep{McNutt.1996}. The extent of
the flood greatly changes within and between years depending on the amount of
rain that descends from the catchment areas in Angola and reaches the distal
ends of the delta between July and August (see \Cref{FloodPulse} in
\Cref{Appendix:FloodmappingAlgorithm} for an illustration of a typical flood
pulse). The flood thereby drastically affects surrounding landscapes, so that
during maximum extent (ca. 12'000 km\textsuperscript{2}) the delta becomes a
patchy conglomerate of swamps, open water, and islands, whereas it shrinks
substantially (ca. 5'000 km\textsuperscript{2}) during dry periods
\citep{Gumbricht.2004, Wolski.2017}. Most of the Okavango Delta's unique basin
is under protection, such that human influence in our core study area was mainly
restricted to the surroundings of Maun and the smaller villages of Khwai,
Mababe, and Sankuyo.

% Large Study Area
The \textit{extended study area} (centered at -17\degree 13'9''S, 23\degree
56'4''E; elevation ca. 1,030 m, red rectangle in \Cref{StudyArea}a) was confined
by a bounding box encompassing the entire KAZA-TFCA. The KAZA-TFCA lies in the
basins of the Okavango and Zambezi rivers and embraces parts of Angola,
Botswana, Namibia, Zimbabwe, and Zambia. With a total area of over 520,000
km\textsuperscript{2} it contains several already-existing national parks (NPs)
and other protected areas. It constitutes the earth's largest transboundary
conservation area, which resulted in a rectangular study area of over 1.3 Mio
km\textsuperscript{2}. Similar to the core study area, the extended study area
hosts various landscapes, including savanna, grassland, and dry or moist
woodland habitats. Despite 36 NPs and other protected areas, there is
considerable human influence in some regions of the KAZA-TFCA originating from
farms, high human density, and road traffic.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.pdf}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[black] (0.202, 0.255) -- (0.36, 0.955);
            \draw[black] (0.202, 0.205) -- (0.36, 0.070);
        \end{scope}
    \end{tikzpicture}
    \caption{Overview of our study area. (a) The red rectangle depicts our
    \textit{extended study area}, which was confined by the extent of the
    KAZA-TFCA. Purple areas indicate remaining wild dog populations according to
    the IUCN \citep{Woodroffe.2012}. (b) The orange rectangle illustrates our
    \textit{core study area}, from which all dispersers departed. Yellow lines
    depict all dispersal trajectories that were recorded between the years 2011
    and 2019. Since game reserves in Botswana virtually serve the same purpose as
    National Parks, we will use the terms interchangeably for this region.}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
% Collection
We collected GPS relocation data of 16 wild dog coalitions (i.e. 7 female and 9
male coalitions) that dispersed from the core study area between 2011 and 2019.
Potential dispersing individuals were immobilized according to protocols
described in \cite{Osofsky.1996} and fitted with GPS/Satellite radio collars
(\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) while they were still
with their natal pack. All required procedures were undertaken and supervised by
a Botswana-registered wildlife veterinarian. Immobilized individuals were
monitored and all were observed to successfully re-join their packs within one
hour after immobilization. GPS collars were programmed to record GPS relocation
data every 4 hours during dispersal and to regularly transmit them via Iridium
satellite system to a base station. The system thus allowed remote-tracking of
collared individuals, even where field conditions were prohibitive or dispersal
trajectories crossed international borders.

% Separation of Dispersers/Residents
Because we were mainly interested in recording dispersal trajectories, we
discarded any GPS data collected while individuals were still with their natal
packs \citep{Cozzi.2020}. Dispersal was deemed to have started when a coalition
had left its natal home range, and continued until it exhibited clear signs of
resident behavior. Some coalitions undertook exploratory movements prior to
proper dispersal. Such movements have been observed in other species and are
reported to be very similar to proper dispersal \citep{Killeen.2014}. We
therefore classified exploratory movements as dispersal too.

% Some Summary Statistics
In total, we collected 4'691 GPS relocations during dispersal (see
\Cref{Trajectories}). In our analysis, we did not differentiate between male and
female dispersing coalitions, for previous research found little differences
between sexes during dispersal \citep{Woodroffe.2019, Cozzi.2020}. On average,
dispersal coalitions included three individuals (min = 1, max = 7), dispersed
for 41 days (min = 7, max = 133), and covered a mean cumulative distance of 370
km (min = 105, max = 610).

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Trajectories.pdf}
    \caption{Illustration of the dispersal trajectories that we recorded. Each
    color represents a different dispersal coalition. All coalitions departed
    from our core study area, which is indicated by the white square. The
    coalition dispersing towards the far east of the map covered over 360 km in
    under 10 days.}
    \label{Trajectories}
  \end{center}
\end{figure}

\subsection{Spatial Covariates}
% General stuff
We used a set of geo-referenced covariates to investigate habitat preferences of
dispersing wild dogs (see \Cref{Covariates}). Covariates fell into one of the
three categories \textit{land cover}, \textit{protection status}, and
\textit{human influence}. Note that we did not include elevation as covariate
due to the absence of noteworthy elevational gradients in our study area. For
each covariate we obtained spatial raster layers from freely available online
services or from remote sensed satellite imagery. To ensure a consistent
resolution (i.e. cell-size or grain) across covariates, all layers were
coarsened or interpolated to match a resolution of 250m x 250m. Processing and
manipulation of data, as well as all spatial and statistical analyses were
performed in R, version 3.6.1 \citep{R.2019}.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Covariates.pdf}
    \caption{Overview of covariates that we used in our models. Although all
    covariates were prepared for the extended study area, we present only an
    extract to maximize visibility. The white square in each plot depicts our
    core study area. (a) Averaged layer of all binary maps of water and dryland
    that we acquired using our floodmapping algorithm. (b) Percentage cover of
    trees. (c) Percentage cover of shrubs/grassland. Note that anything not
    covered by trees or shrubs/grassland was deemed to be bareland. (d)
    Protection status according to the Peace Parks Foundation. (e) Human
    influence proxy, including human density, farms, and roads. (f) Distance to
    nearest road (white lines depict actual roads).}
    \label{Covariates}
  \end{center}
\end{figure}

\newpage
\subsubsection{Land Cover}
% Overview
Land cover was represented by four layers: a binary layer distinguishing water
and dryland, and three continuous layers showing the percentage cover by
different vegetation types, together adding up to 100\%.

% Water Layers
Water included open waters, rivers, wetlands, and swamps, whereas anything else
was presumed to be dryland. Because the flood extent of the Okavango Delta is
highly variable within and between years, we created weekly (8 days)
``floodmaps'', following the algorithm developed in \cite{Wolski.2017}. The
algorithm uses thresholding of SWIR reflectances of MODIS Terra satellite
imagery (MCD43A4, \cite{Schaaf.2015}) to separate water and dryland. As we
rewrote and slightly amended the original Python code in R, we describe the
underlying principles and indicate the discrepancies to the original algorithm
in \Cref{Appendix:FloodmappingAlgorithm}. While water in the Okavango Delta was
mapped dynamically, we used a static watermap for its surroundings. This static
map was based on Globeland's land cover dataset \citep{Chen.2015}, from which we
only retained the categories \textit{wetland} and \textit{waterbodies} and
collectively reclassified them to \textit{water}. Globeland had an original
resolution of 30m x 30m, so we coarsened the layer to 250m x 250m using the mode
of each 250m x 250m cell. Lastly, we improved river representation using the
MERIT Hydro dataset \citep{Yamazaki.2019}, from which we added all rivers with a
width of over 10m to our watermaps.

>>>>

% Title of the document
{\textsc{ \LARGE On the Move: Dispersal Corridors of African Wild Dogs
(\textit{Lycaon pictus}) in the Kavango-Zambezi Transfrontier Conservation
Area}}\\[0.4cm]

@@

\begin{abstract}
The African wild dog \textit{(Lycaon pictus)} is an endangered species that
currently only survives in a few scattered subpopulations in Sub-Saharan Africa.
Long-term viability of the species thus requires to identify and preserve key
dispersal corridors that connect such subpopulations. We collected GPS data of
16 dispersing coalitions originating from a free-ranging population in northern
Botswana and used integrated step selection analysis to create a permeability
surface spanning over the entire Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA), the largest transboundary conservation area on earth. Based on our
permeability surface, we calculated least-cost paths and least-cost corridors
between selected locations within the KAZA-TFCA. Our results show that dispersal
corridors typically run through open shrubs/grassland and in the vicinity of
water. Corridors rarely cross water bodies, areas with high human presence, or
areas densely vegetated by trees. Several dispersal corridors meet in the
Okavango-Linyanty ecosystem, which highlights the importance of northern
Botswana for the conservation of this charismatic species.
\end{abstract}
\newpage

@@

\section{Introduction}
% Importance of Connectivity
Successful conservation of fragmented and isolated subpopulations requires to
maintain and enhance connectivity between habitats occupied by these
subpopulations \citep{Hanski.1998}. In this context, connectivity describes the
degree to which the landscape matrix facilitates or impedes movements of
individuals of the focal species between habitat patches \citep{Taylor.1993,
Clobert.2012}. Thus, sufficient connectivity facilitates dispersal, which in
turn promotes genetic exchange, reinforcement and rescuing of small populations,
and colonization or recolonization of unoccupied habitats \citep{Brown.1977,
Hanski.1998, MacArthur.2001, Frankham.2002, Leigh.2012}. These processes
significantly contribute to long-term population viability, which emphasizes the
importance of connectivity. However, increasing anthropogenic pressure and
ongoing habitat loss and fragmentation reduce connectivity on a worldwide scale
\citep{Fahrig.2003, Barnosky.2012}. As a result, management actions for the
conservation of endangered species increasingly focus on preserving and
restoring connectivity through the identification and preservation of key
dispersal corridors that link isolated habitats \citep{Fahrig.2003,
Heller.2009}. An unprecedented initiative to restore connectivity on a large
scale is represented by the creation of the Kavango-Zambezi Transfrontier
Conservation Area (KAZA-TFCA), a conservation area that traverses five African
countries and spans over 520,000 km\textsuperscript{2}. While such initiatives
are expected to bring benefits to wide-ranging species such as elephants
(\textit{Loxodonta africana}), African wild dogs (\textit{Lycaon pictus}), and
lions (\textit{Panthera leo}), the extent to which they are effective for
targeted species and whether boundaries reflect natural dispersal corridors is
rarely assessed.

% Connectivity Assessment
In recent years, there has been a growing body of literature that uses animal
relocation data to identify dispersal corridors and to assess connectivity on
large scales \citep{Chetkiewicz.2006}. Identification of potential dispersal
corridors between suitable habitat patches traditionally relies on the
estimation of a permeability surface (elsewhere referred to as resistance
surface, when inverted), which renders the ease or willingness at which the
focal species traverses a specific landscape. Latest empirical methods to
predict landscape permeability based on recorded GPS relocation data include
\textit{resource selection functions} (RSF, \cite{Boyce.2002}), \textit{step
selection functions} (SSF, \cite{Fortin.2005}), and \textit{path selection
functions} (PSF, \cite{Cushman.2010}). All methods have in common that they
infer habitat preferences of the focal species by comparing spatial covariates
(e.g. environmental and anthropogenic) at visited locations to the same spatial
covariates at randomly selected locations that are deemed available but
presumably unused by the animal. Based on inferred habitat preferences the
methods further assume that habitats preferred by the focal species provide high
permeability, whereas avoided habitats provide low permeability. Regardless of
the method chosen (i.e. RSF, SSF, or PSF), all approaches require adequate GPS
relocation data that is representative of the process being studied. More
specifically, it has been shown that GPS relocation data collected on dispersing
individuals are better suited for detecting dispersal corridors than GPS
relocation data collected on resident individuals \citep{Elliot.2014}. Few
studies have, however, partitioned their GPS relocation data according to
behavioral modes such as \textit{resident} vs. \textit{dispersing} or
\textit{active} vs. \textit{inactive} periods of movement \citep{Wilson.2012,
Vasudev.2015}. As a consequence, most permeability surfaces upon which dispersal
corridors are identified are created using GPS relocation data that was
collected on resident individuals. This introduces severe biases and
substantially reduces the power to reveal meaningful corridors, for dispersing
individuals have different needs and motivations compared to resident
individuals (\citeauthor{Killeen.2014}, 2014; \citeauthor{Elliot.2014}, 2014;
\citeauthor{Cozzi.2020}, unpublished). For instance, \cite{Elliot.2014}
demonstrated marked differences in predicted permeability surfaces for
dispersing lions in the KAZA-TFCA depending on whether permeability was assessed
using data collected during residence or dispersal. \cite{Jackson.2016} made
similar observations for African wild dogs and revealed higher permeability for
dispersers than for residents. Moreover, \cite{Abrahms.2017} found a
significantly greater overlap between predicted and observed corridors for three
dispersing wild dogs when predictions were based on information collected only
during wild dog activity, rather than across the entire day.

% African wild dogs
A highly endangered flagship species for conservation efforts across the
KAZA-TFCA, but for which systematic GPS relocation data on dispersing
individuals is largely unavailable, is the African wild dog
\citep{VanDerMeer.2016}. While once present across the entire Sub-Saharan
continent, the species has been widely extirpated through human persecution,
habitat destruction, and disease outbreaks \citep{Woodroffe.2012}. As a result,
the African wild dog has become one of Africa's most endangered large carnivores
and currently only survives in small, spatially scattered subpopulations
\citep{Woodroffe.2012}. Within these subpopulations, wild dogs form cooperative
breeding groups of up to thirty individuals among which a single dominant pair
monopolizes the majority of reproduction \citep{Frame.1979, Fuller.1992,
Creel.2002}, while subordinates of both sexes help to feed newborn pups
\citep{Malcolm.1982}. After reaching 13 months of age, male and female offspring
disperse from their natal pack in same-sex coalitions and search for unrelated
mates and a suitable territory to settle (\citeauthor{McNutt.1996}, 1996;
\citeauthor{Behr.2019}, unpublished). During such dispersal events, wild dogs
can cover several hundred kilometers, sometimes crossing risky areas that are
strongly influenced by humans (\citeauthor{DaviesMostert.2012}, 2012;
\citeauthor{Masenga.2016}, 2016; \citeauthor{Cozzi.2020}, unpublished). Despite
the importance of dispersal for the remaining wild dog subpopulations, little is
known about the corridors of dispersing individuals and connectivity between
subpopulations in the KAZA-TFCA in general.

% Covariate Motivation
Wild dog movements and spatial behavior are known to be influenced by a suite of
environmental and anthropogenic factors. For instance, \cite{Abrahms.2017} found
that wild dog movements during residence mainly depended on land cover,
protection status, and anthropogenic features. Similar observations on resident
packs were made by \cite{Pretorius.2019}, who's results indicate that pack
presence was contingent on factors such as agricultural activity, road presence,
and protection status of the area. Concerning dispersing individuals,
\cite{Whittington.2011} demonstrated that elevation, land cover, road density,
and human density best explained the distribution of recorded sightings in South
Africa. Also based on data from dispersing wild dogs, \cite{Jackson.2016} showed
that land cover, human density, and road density significantly influenced the
probability of occurrence of dispersing wild dogs. Finally, \cite{Masenga.2016}
reported a low willingness of dispersers to cross human-dominated landscapes.
Large scale spatial information on the above-mentioned factors can be obtained
either through freely available online services or through remotely sensed
satellite imagery.

% Our Study
We used GPS relocation data collected on 16 wild dogs in as many dispersing
coalitions from a free-ranging population in northern Botswana to estimate the
degree of selection or avoidance for environmental and anthropogenic landscape
features. We then used the obtained information in combination with relevant
spatial covariates to create a permeability surface for the entire KAZA-TFCA.
Lastly, we used the permeability surface as input for the computation of
least-cost paths and least-cost corridors between selected locations. This
allowed us to highlight key dispersal corridors within and outside the
boundaries of the KAZA-TFCA and to identify a potential ``dispersal hub'' as
well as other dispersal corridors that run outside formally protected areas.

\newpage
\section{Methods}
\subsection{Study Area}
% Both Study Areas
We differentiated between a \textit{core study area}, where all dispersers
originated from, and an \textit{extended study area}, which was represented by a
bounding box encompassing the KAZA-TFCA (\citeauthor{Cozzi.2020}, unpublished)
and for which we calculated connectivity.

% Core Study Area
The \textit{core study area} was located in the eastern section of the Okavango
Delta in northern Botswana (centered at -19\degree 25'30''S, 23\degree 37'30''E;
elevation ca. 950 m, orange rectangle in \Cref{StudyArea}b) and comprised parts
of the Moremi Game Reserve and adjacent wildlife management areas. The Okavango
Delta is one of the main strongholds of African wild dogs in southern Africa and
likely acts as a source for the recolonization of surrounding habitats
\citep{Woodroffe.2012, Cozzi.2013}. It is the earth's largest inland delta and
is characterized by diverse habitats, ranging from permanent river systems,
mopane woodland, and acacia forests, to open grasslands and seasonal swamps
\citep{Broekhuis.2013}. Rainfall is seasonal and lasts from November to March,
whereas the remainder of the year remains dry \citep{McNutt.1996}. The extent of
the flood greatly changes within and between years depending on the amount of
rain that descends from the catchment areas in Angola and reaches the distal
ends of the delta between July and August (see \Cref{FloodPulse} in
\Cref{Appendix:FloodmappingAlgorithm} for an illustration of a typical flood
pulse). The flood drastically affects surrounding landscapes, so that during
maximum extent (ca. 12'000 km\textsuperscript{2}) the delta becomes a patchy
conglomerate of swamps, open water, and islands, whereas these structures
disappear when the flood retracts (ca. 5'000 km\textsuperscript{2}) during dry
periods \citep{Gumbricht.2004, Wolski.2017}. Most of the Okavango Delta's unique
basin is under protection, such that human influence in the core study area was
mainly restricted to the surroundings of Maun and the smaller villages of Khwai,
Mababe, and Sankuyo.

% Large Study Area
The \textit{extended study area} (centered at -17\degree 13'9''S, 23\degree
56'4''E; elevation ca. 1,030 m, red rectangle in \Cref{StudyArea}a) was confined
by a bounding box encompassing the entire KAZA-TFCA. The KAZA-TFCA lies in the
basins of the Okavango and Zambezi rivers and embraces parts of Angola,
Botswana, Namibia, Zimbabwe, and Zambia. With a total area of over 520,000
km\textsuperscript{2}, it contains several already-existing national parks (NPs)
and other protected areas. It constitutes the earth's largest transboundary
conservation area, which resulted in a rectangular study area of over 1.3 Mio
km\textsuperscript{2}. Similar to the core study area, the extended study area
hosts various landscapes, including savanna, grassland, and dry or moist
woodland habitats. Despite 36 NPs and other protected areas, there is
considerable human influence in some regions of the KAZA-TFCA originating from
farms, high human density, and road traffic.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=0.9\textwidth]{99_StudyArea2.pdf}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[black] (0.202, 0.255) -- (0.36, 0.955);
            \draw[black] (0.202, 0.205) -- (0.36, 0.070);
        \end{scope}
    \end{tikzpicture}
    \caption{Overview of our study area. (a) The red rectangle depicts our
    \textit{extended study area}, which was confined by the extent of the
    KAZA-TFCA. Purple areas indicate remaining wild dog populations according to
    the IUCN \citep{Woodroffe.2012}. (b) The orange rectangle illustrates our
    \textit{core study area}, from which all dispersers departed. Yellow lines
    depict all dispersal trajectories that were recorded between the years 2011
    and 2019. Since game reserves in Botswana virtually serve the same purpose as
    National Parks, we will use the terms interchangeably for this region.}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
% Collection
We collected GPS relocation data of 16 wild dog coalitions (7 female and 9
male coalitions) that dispersed from the core study area between 2011 and 2019
(see also \citeauthor{Abrahms.2017}, 2017 and \citeauthor{Cozzi.2020},
unpublished). Potential dispersing individuals were immobilized according to
protocols described in \cite{Osofsky.1996} and fitted with GPS/Satellite radio
collars (\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) while they were
still with their natal pack. All required procedures were undertaken and
supervised by a Botswana-registered wildlife veterinarian. Immobilized
individuals were monitored and all were observed to successfully re-join their
packs within one hour after immobilization. GPS collars were programmed to
record GPS relocation data every 4 hours during dispersal and to regularly
transmit them via Iridium satellite system to a base station. The system thus
allowed remote-tracking of collared individuals, even where field conditions
were prohibitive or dispersal trajectories crossed international borders.

% Separation of Dispersers/Residents
Because we were mainly interested in recording dispersal trajectories, we
discarded any GPS data collected while individuals were still with their natal
packs (\citeauthor{Cozzi.2020}, unpublished). Dispersal was deemed to have
started when a coalition had left its natal home range, and continued until it
exhibited clear signs of resident behavior. Some coalitions undertook
exploratory movements prior to proper dispersal. Such movements have been
observed in other species and are reported to be very similar to proper
dispersal \citep{Killeen.2014}. We therefore classified exploratory movements as
dispersal too.

% Some Summary Statistics
In total, we collected 4'691 GPS relocations during dispersal (see
\Cref{Trajectories}). In our analysis, we did not differentiate between male and
female dispersing coalitions, for previous research found little differences
between sexes during dispersal (\citeauthor{Woodroffe.2019}, 2019;
\citeauthor{Cozzi.2020}, unpublished). On average, dispersal coalitions included
three individuals (min = 1, max = 7), dispersed for 41 days (min = 7, max =
133), and covered a mean cumulative distance of 370 km (min = 105, max = 610).

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Trajectories2.pdf}
    \caption{Illustration of the dispersal trajectories that we recorded. Each
    color represents a different dispersal coalition. All coalitions departed
    from the core study area, which is indicated by the white rectangle. The
    coalition dispersing towards the far east of the map covered over 360 km in
    under 10 days.}
    \label{Trajectories}
  \end{center}
\end{figure}

\subsection{Spatial Covariates}
% General stuff
We used a set of geo-referenced covariates to investigate habitat preferences of
dispersing wild dogs (see \Cref{Covariates}). Covariates fell into one of the
three categories of \textit{land cover}, \textit{protection status}, and
\textit{human influence}. We did not include elevation as a covariate due to the
absence of noteworthy elevational gradients in our study area. For each selected
covariate, we obtained spatial raster layers from freely available online
services or remotely sensed satellite images. To ensure a consistent resolution
(i.e. cell-size or grain) across covariates, all layers were coarsened or
interpolated to match a resolution of 250m x 250m. Processing and manipulation
of data, as well as all spatial and statistical analyses, were performed in R,
version 3.6.1 \citep{R.2019}.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width = \textwidth]{99_Covariates.pdf}
    \caption{Overview of covariates that we used in our models. Although all
    covariates were prepared for the extended study area, we present only an
    extract to maximize visibility. The white rectangle in each plot depicts our
    core study area. (a) Averaged layer of all binary maps of water and dryland
    that we acquired using our floodmapping algorithm. (b) Percentage cover of
    trees. (c) Percentage cover of shrubs/grassland. Note that anything not
    covered by trees or shrubs/grassland was deemed to be bare land. (d)
    Protection status of the area. (e) Human influence proxy, taking into
    account human density, farms, and roads. (f) Distance to nearest road (white
    lines depict actual roads).}
    \label{Covariates}
  \end{center}
\end{figure}

\newpage
\subsubsection{Land Cover}
% Overview
Land cover was represented by four layers: a binary layer distinguishing water
and dryland, and three continuous layers showing the percentage cover by
different vegetation types, combined adding up to 100\%.

% Water Layers
Water included open waters, rivers, wetlands, and swamps, whereas anything else
was presumed to be dryland. Because the flood extent of the Okavango Delta is
highly variable within and between years, we created weekly (8 days)
``floodmaps'', following the algorithm developed in \cite{Wolski.2017}. The
algorithm uses thresholding of SWIR reflectances of MODIS Terra satellite
imagery (MCD43A4, \cite{Schaaf.2015}) to separate water and dryland. As we
rewrote and slightly amended the original Python code in R, we describe the
underlying principles and indicate the discrepancies to the original algorithm
in \Cref{Appendix:FloodmappingAlgorithm}. While water in the Okavango Delta was
mapped dynamically, we used a static watermap for its surroundings. This static
map was based on Globeland's land cover dataset \citep{Chen.2015}, from which we
only retained the categories \textit{wetland} and \textit{water bodies} and
collectively reclassified them to \textit{water}. Globeland had an original
resolution of 30m x 30m, so we coarsened the layer to 250m x 250m using the mode
of each 250m x 250m cell. Lastly, we improved river representation using the
MERIT Hydro dataset \citep{Yamazaki.2019}, from which we added all rivers with a
width of over 10m to our floodmaps.
